\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2024-11-25 main.tex
change to typed lambda calculus

% \section{About Typing}
% I plan to implement a typing system in C++, replicating the one in Coq.

>>>>

\clearpage
\section{Typing System}
The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
According to the CoC typing rules, the environment and context can be modeled by as two stacks. It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.

One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.

\newcommand{\reduce}{\triangleright}

\newcommand{\Sort}{\mathsf{Sort}}
\newcommand{\WF}{\mathcal{WF}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\BaseS}{\mathsf{Base}}

\newcommand{\SType}{\mathcal{S}}
\newcommand{\KType}{\mathcal{K}}
\newcommand{\BType}{\mathcal{B}}
\newcommand{\OType}{\mathcal{O}}

\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}

\begin{figure}[h]
    \begin{gather*}
        \textbf{W-Empty} \qquad
        \frac{}{\WF([])}\\
        \\
        \textbf{W-Assum} \qquad
        \frac{E\vdash T:\Type \qquad c \notin E}{\WF(E; c:T)}
        \qquad \qquad
        \textbf{W-Def} \qquad
        \frac{E\vdash t:T \qquad c \notin E}{\WF(E; c:=t:T)} \\
        \\
        \textbf{Const} \qquad
        \frac{\WF(E) \qquad (c:T) \in E \text{ or } (c := t : T) \in E \text{ for some $t$}}{E \vdash c : T} \\
        \\
        \textbf{Arrow} \qquad
        \frac{E \vdash T:\Type \qquad E \vdash U:\Type}{E \vdash T \to U : \Type} \\
        \\
        \textbf{Lam} \qquad
        \frac{E;x:T \vdash t : U}{E \vdash (\lambda x:T . t) : T \to U}
        \qquad \qquad
        \textbf{App} \qquad
        \frac{E \vdash t:U \to T \qquad E \vdash u:U}{E \vdash (t\ u):T}
    \end{gather*}
    \caption{The typing rules for the typed lambda calculus with environment.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Type-Base} \qquad
        \frac{\WF(E)}{E \vdash \BaseS : \Type} 
        \qquad \qquad
        \textbf{Base-Subtype} \qquad
        \frac{E \vdash T : \BaseS}{E \vdash T : \Type} \\
        \\
        \textbf{Type-Ket} \qquad
        \frac{E \vdash \sigma : \BaseS}{E \vdash \KType(\sigma) : \Type}
        \qquad \qquad
        \textbf{Type-Bra} \qquad
        \frac{E \vdash \sigma : \BaseS}{E \vdash \BType(\sigma) : \Type} \\
        \\
        \textbf{Type-Opt} \qquad
        \frac{E \vdash \sigma : \BaseS \qquad E \vdash \tau : \BaseS}{E \vdash \OType(\sigma, \tau) : \Type}
        \qquad \qquad
        \textbf{Type-Scalar} \qquad
        \frac{\WF(E)}{E \vdash \SType : \Type}
    \end{gather*}
    \caption{Typing rules for Dirac notation types.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Sca-0} \qquad
        \frac{\WF(E)}{E \vdash 0 : \SType}
        \qquad \qquad
        \textbf{Sca-1} \qquad
        \frac{\WF(E)}{E \vdash 1 : \SType}
        \qquad \qquad
        \textbf{Sca-Delta} \qquad
        \frac{            
            \begin{aligned}
                & E\vdash \sigma : \BaseS \\
                & E\vdash s : \sigma \qquad E\vdash t : \sigma
            \end{aligned}} {E \vdash \delta_{s, t} : \SType} \\
        \\
        \textbf{Sca-Add} \qquad
        \frac{E\vdash a : \SType \qquad E\vdash b : \SType}{E\vdash a + b : \SType}
        \qquad \qquad
        \textbf{Sca-Mul} \qquad
        \frac{E\vdash a : \SType \qquad E\vdash b : \SType}{E\vdash a \times b : \SType} \\
        \\
        \textbf{Sca-Conj} \qquad
        \frac{E \vdash a : \SType}{E \vdash a^*:\SType}
        \qquad \qquad
        \textbf{Sca-Dot} \qquad
        \frac{E\vdash B : \BType(\sigma) \qquad E\vdash K : \KType(\sigma)}{E \vdash B \cdot K : \SType}
    \end{gather*}
    \caption{Scalar typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Base-Def} \qquad
        \frac{\WF(E) \qquad a_i \notin E \text{ for all $i$}}
        {E\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS}
        \qquad \qquad
        \textbf{Base-Element} \qquad
        \frac{\WF(E) \qquad c \in \{a_1, a_2, \cdots a_n\}}{E\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
        \\
        \textbf{Base-Prod} \qquad
        \frac{E\vdash \sigma : \BaseS \qquad E\vdash \tau : \BaseS}{E\vdash \sigma \times \tau : \BaseS}
        \qquad \qquad
        \textbf{Pair-Base} \qquad
        \frac{
            \begin{aligned}
                & E\vdash \sigma : \BaseS \qquad E\vdash s : \sigma \\
                & E\vdash \tau : \BaseS \qquad E\vdash t : \tau
            \end{aligned}} {E\vdash (s, t) : \sigma \times \tau} 
        \textbf{}
    \end{gather*}
    \caption{Typing rules for bases. }
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Ket-0} \qquad
        \frac{E \vdash \sigma : \BaseS}{E \vdash \ZEROK(\sigma) : \KType(\sigma)} 
        \qquad \qquad
        \textbf{Ket-Base} \qquad
        \frac{E \vdash \sigma : \BaseS \qquad E\vdash t : \sigma}{E \vdash \ket{t} : \KType(\sigma)} \\
        \\
        \textbf{Ket-Adj} \qquad
        \frac{E \vdash B : \BType(\sigma)}{E \vdash B^\dagger : \KType(\sigma)} 
        \qquad \qquad
        \textbf{Ket-Sca} \qquad
        \frac{E \vdash a : \SType \qquad E \vdash K : \KType(\sigma)}{E \vdash a.K : \KType(\sigma)} \\
        \\
        \textbf{Ket-Add} \qquad
        \frac{E \vdash K_1 : \KType(\sigma) \qquad E \vdash K_2 : \KType(\sigma)}{E \vdash K_1 + K_2 : \KType(\sigma)}
        \qquad \qquad
        \textbf{Ket-Dot} \qquad
        \frac{E \vdash O : \OType(\sigma, \tau) \qquad E \vdash K : \KType(\tau)}{E \vdash O \cdot K : \KType(\sigma)} \\
        \\
        \textbf{Ket-Tsr} \qquad
        \frac{E \vdash K_1 : \KType(\sigma) \qquad E \vdash K_2 : \KType(\tau)} {E \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
    \end{gather*}
    \caption{Ket typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Bra-0} \qquad
        \frac{E \vdash \sigma : \BaseS}{E \vdash \ZEROB(\sigma) : \BType(\sigma)} 
        \qquad \qquad
        \textbf{Bra-Base} \qquad
        \frac{E \vdash \sigma : \BaseS \qquad E\vdash t : \sigma}{E \vdash \bra{t} : \BType(\sigma)} \\
        \\
        \textbf{Bra-Adj} \qquad
        \frac{E \vdash K : \KType(\sigma)}{E \vdash K^\dagger : \BType(\sigma)} 
        \qquad \qquad
        \textbf{Bra-Sca} \qquad
        \frac{E \vdash a : \SType \qquad E \vdash B : \BType(\sigma)}{E \vdash a.B : \BType(\sigma)} \\
        \\
        \textbf{Bra-Add} \qquad
        \frac{E \vdash B_1 : \BType(\sigma) \qquad E \vdash B_2 : \BType(\sigma)}{E \vdash B_1 + B_2 : \BType(\sigma)}
        \qquad \qquad
        \textbf{Bra-Dot} \qquad
        \frac{E \vdash B : \KType(\sigma) \qquad E \vdash O : \OType(\sigma, \tau)}{E \vdash B \cdot O : \BType(\tau)} \\
        \\
        \textbf{Bra-Tsr} \qquad
        \frac{E \vdash B_1 : \BType(\sigma) \qquad E \vdash B_2 : \BType(\tau)} {E \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
    \end{gather*}
    \caption{Bra typing rules.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \textbf{Opt-0} \qquad
        \frac{E \vdash \sigma : \BaseS \qquad E \vdash \tau : \BaseS}{E \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
        \qquad \qquad
        \textbf{Opt-1} \qquad
        \frac{E \vdash \sigma : \BaseS}{E \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
        \\
        \textbf{Opt-Adj} \qquad
        \frac{E \vdash O : \OType(\sigma, \tau)}{E \vdash O^\dagger : \OType(\tau, \sigma)} 
        \qquad \qquad
        \textbf{Opt-Sca} \qquad
        \frac{E \vdash a : \SType \qquad E \vdash O : \OType(\sigma, \tau)}{E \vdash a.O : \OType(\sigma, \tau)} \\
        \\
        \textbf{Opt-Add} \qquad
        \frac{E \vdash O_1 : \OType(\sigma, \tau) \qquad E \vdash O_2 : \OType(\sigma, \tau)}{E \vdash O_1 + O_2 : \OType(\sigma, \tau)}
        \qquad \qquad
        \textbf{Opt-Outer} \qquad
        \frac{E\vdash K : \KType(\sigma) \qquad E\vdash B : \BType(\tau)}{E\vdash K \cdot B : \OType(\sigma, \tau)} \\
        \\
        \textbf{Opt-Dot} \qquad
        \frac{E \vdash O_1 : \OType(\sigma, \tau) \qquad E \vdash O_2 : \OType(\tau, \rho)}{E \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
        \\
        \textbf{Opt-Tsr} \qquad
        \frac{E \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E \vdash O_2 : \OType(\sigma_2, \tau_2)} {E \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
    \end{gather*}
    \caption{Operator typing rules.}
\end{figure}

\clearpage
\section{Conversions and Reductions}
\begin{figure}[h]
    \begin{gather*}
        \textbf{Alpha} \qquad \text{(bound variable renaming)}
        \qquad \qquad
        \textbf{Beta} \qquad
        \frac{}{E \vdash ((\lambda x:T.t)\ u) \reduce_\beta t\{x/u\}} \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E) \qquad (c:=t:T) \in E}{E \vdash c \reduce_\delta t}
        \qquad \qquad
        \textbf{Eta} \qquad
        \frac{E\vdash t : T \to U}{E \vdash \lambda x : T . (t\ x) \reduce_\eta t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

\begin{figure}[h]
    \begin{gather*}
        \cdots
    \end{gather*}
    \caption{Dirac notation reduction rules}
\end{figure}

About big operators: maybe we should use the "sum operators"?

\clearpage
